---
title: "Course Project 1"
author: "Filippo D'Ercoli"
date: "6/17/2020"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lattice)
```

## Loading and preprocessing the data

First, I need to read the data into RStudio. A summary of the dataset is reported.

```{r data overview}
data <- read.csv("activity.csv")
summary(data)
```

## What is mean total number of steps taken per day?

I summed the steps taken in each interval of a single day using "tapply". Then I plotted an histogram and calculated mean and median of the total number of steps taken per day.

```{r}
totalStepsPerDay <- tapply(data$steps, data$date, sum, na.rm=TRUE)
hist(totalStepsPerDay, main = "Total Steps per Day")
summary(totalStepsPerDay) [4]
summary(totalStepsPerDay) [3]
```

## What is the average daily activity pattern?

I grouped the observations by interval and I calculated the mean of each group, obtaining a daily activity pattern. Then I plotted this pattern versus the intervals.

```{r}
dailyPattern <- group_by(data, interval) %>% summarise(intervalStepAvg = mean(steps, na.rm=TRUE))
with(dailyPattern, plot(interval, intervalStepAvg, type = "l"))
title(main = "Daily Activity Pattern")
maxSteps <- max(dailyPattern$intervalStepAvg)
maxInt <- subset(dailyPattern, dailyPattern$intervalStepAvg == maxSteps) [1]
```

The interval which on average registers the maximum number of steps is `r maxInt`.

## Imputing missing values

```{r}
totalNA <- sum(is.na(data$steps))
```

The total number of missing values is `r totalNA`. A suitable strategy for fill them in could be using the mean over the same 5-minute interval. A new dataset "imputedData" with the imputed values was created. Then I plotted the histogram of the new dataset and calculated again mean and median.

```{r}
imputedData <- data
for (i in 1:length(data$steps)) {
  if (is.na(data$steps [i])) {
    if (i%%288 != 0) {
      imputedData$steps [i] <- dailyPattern$intervalStepAvg [i%%288]
    }
    else imputedData$steps [i] <- dailyPattern$intervalStepAvg [288]
  }
}
totalStepsPerDay2 <- tapply(imputedData$steps, imputedData$date, sum, na.rm=TRUE)
hist(totalStepsPerDay2, main = "Total Steps per Day")
summary(totalStepsPerDay2) [4]
summary(totalStepsPerDay2) [3]
```

Imputing the data gives the histogram a more regular shape. The median is almost the same, while the mean increases a bit and happens to be equal to the median.

## Are there differences in activity patterns between weekdays and weekends?

I created a new variable called "dayLevel". It is a factor variable with 2 levels which identify weekdays and weekends.

```{r}
weekday <- weekdays(as.POSIXct.Date(imputedData$date))
weekData <- mutate(imputedData, dayLevel = rep("weekday", length(imputedData$date)))
for (i in 1:length(imputedData$date)) {
  if (weekday[i] == "Saturday" || weekday[i] == "Sunday") {
    weekData$dayLevel[i] <- "weekend"
  }
}
```

Then I grouped the observations by interval and dayLevel, calculated the average steps taken per interval and plotted it.

```{r}
weeklyPattern <- group_by(weekData, interval, dayLevel) %>% summarise(intervalStepAvg = mean(steps, na.rm=TRUE))
xyplot(intervalStepAvg ~ interval | dayLevel, weeklyPattern, type = "l")
```
